\singlespacing \vspace{-1.6cm} \LARGE Trade-offs among plant reproductive traits determine interactions with floral visitors

### 2022-01-26 Manuscript/Lanuza_et_al_2021_Ecol_Letters.tex
edits

\textbf{Plant-pollinator network studies.} We selected 28 studies from 18 different countries that constituted a total of 64 plant-pollinator networks. These studies recorded plant-pollinator interactions in natural systems and were selected so that we had broad geographical representation. Although these studies differ in sampling effort and methodology, all studies provided information about plant-pollinator interactions (weighted and non-weighted), which we used to build a database of plant species that are likely to be animal pollinated. Many of these networks are freely available either as published studies (\protect\hyperlink{ref-olesen2007}{Olesen \emph{et al.} 2007}; \protect\hyperlink{ref-fortuna2010}{Fortuna \emph{et al.} 2010}; \protect\hyperlink{ref-carvalheiro2014}{Carvalheiro \emph{et al.} 2014}) or available in online archives (e.g., The Web of Life, \protect\hyperlink{ref-fortuna2010}{Fortuna \emph{et al.} 2010}; and Mangal, \protect\hyperlink{ref-poisot2016}{Poisot \emph{et al.} 2016}). In total, our network dataset (see Table S1) constituted 60 weighted (interaction frequency) and 4 unweighted (presence/absence of the interaction) networks, each sampled at a unique location and year, as well as eight meta-webs where interactions were pooled across several locations and multiple years.

\textbf{Taxonomy of plants and pollinators.} All species names, genera, families and orders were retrieved and standardized from the taxonomy data sources NCBI (\url{https://www.ncbi.nlm.nih.gov/taxonomy}) for plants and ITIS (\url{https://www.itis.gov/}) for pollinators, using the R package \emph{taxize} (\protect\hyperlink{ref-chamberlain2020}{Chamberlain \emph{et al.} n.d.b}) version \emph{0.9.99}. We filled the `not found' searches manually using \url{http://www.theplantlist.org/} and \url{http://www.mobot.org/} for plants and \url{http://www.catalogueoflife.org/} for floral visitors.

\textbf{Functional traits.} We selected a total of 19 different functional traits that included both reproductive and vegetative traits (see Table 1 and Supplementary Information). From these, 16 were reproductive traits (13 floral and 3 reproductive biology traits) and were selected based on their relevance to plant reproduction and data availability. The 3 remaining traits, were vegetative traits that are commonly used to characterize the global spectrum of plant form and function and represent the fast-slow continuum of trait variation (e.g., short-lived versus perennial species). For each plant species, we undertook an extensive literature and online search across a wide range of resources (plant databases, online floras, books, journals and images). From a total of 30,120 cells (20 columns × 1,506 species) we were able to fill 24,341 cells (80.8\% of the dataset, see Fig. S1 for missing values information for each trait).

\textbf{Phylogenetic Distance.} We calculated the phylogenetic distance between different plant species using the function \emph{get\_tree} from the package \emph{rtrees} (\url{https://github.com/daijiang/rtrees}), which downloads phylogenetic distances from the extended R implementation of the Open Tree of Life (\protect\hyperlink{ref-smith2018}{Smith \& Brown 2018}; \protect\hyperlink{ref-jin2019}{Jin \& Qian 2019}).

@@

\textbf{Data Imputation.} Trait missing values were imputed with the function \emph{missForest} (\protect\hyperlink{ref-stekhoven2012}{Stekhoven \& Bühlmann 2012}) which allows imputation of data sets with continuous and categorical variables. We accounted for the phylogenetic distance among species on the imputation process by including the eigenvectors of a principal component analysis of the phylogenetic distance (PCoA) which has been shown to improve the performance of \emph{missForest} (\protect\hyperlink{ref-penone2014}{Penone \emph{et al.} 2014}). To extract the eigenvectors, we used the function \emph{PVRdecomp} from the package \emph{PVR} (\protect\hyperlink{ref-santos2018}{Chamberlain \emph{et al.} n.d.a}) based on a previous conceptual framework that considers phylogenetic eigenvectors (\protect\hyperlink{ref-diniz-filho2012}{Diniz-Filho \emph{et al.} 2012}). Although the variable of autonomous selfing had a high percentage of missing values (68\%), we were able to solve this by back transforming the qualitative column of autonomous selfing to numerical. The categories of `none,' `low,' `medium' and `high' were converted to representative percentages of each category 0\%, 13\%, 50.5\% and 88\% respectively. This reduced the percentage of missing values for this column from 68\% to 35\% and allowed the imputation of this variable. However, we were unable to include nectar and pollen traits on the imputation process because of the high percentage of missing values (Supplementary Fig. S1). Hence, the imputed dataset had 1,506 species, seven categorical and eight numerical variables and 5.79\% of missing values. Further, we conducted an additional imputation process on the subset of species with data for pollen per flower and microliters of nectar. This subset comprised 755 species, 8.01\% missing values and all traits but milligrams of nectar (\textasciitilde50\% of missing values) were included in the imputation process.

\textbf{Plant strategies.} We explored the trade-offs between different quantitative plant functional traits with a phylogenetically informed Principal Component Analysis (pPCA). We did not include the quantitative variables of flower length and inflorescence width because they were highly and moderately correlated to flower width respectively (Pearson's correlation = 0.72, \emph{P} \textless{} 0.01 and Pearson's correlation = 0.36, \emph{P} \textless{} 0.01), and thus we avoided overemphasizing flower size on the spectrum of trait variation. Although qualitative traits were not included in the dimensionality reduction analysis, we also investigated the association of the different qualitative traits with the main axes of trait variation. Prior to the analyses, we excluded outliers and standardized the data. Due to the high sensitivity of dimensionality reduction to outliers, we excluded values within the 2.5th--97.5th percentile range (\protect\hyperlink{ref-legendre2012}{Legendre \& Legendre 2012}), and thus our final dataset had 1,236 species. Then, we log transformed the variables to reduce the influence of outliers and z-transformed (X= 0, SD=1) so that all variables were within the same numerical range. We performed the pPCA using the function \emph{phyl.pca} from the package \emph{phytools} (\protect\hyperlink{ref-revell2012}{Revell 2012}) (version \emph{0.7-70}) with the method lambda (\(\lambda\)) that calculates the phylogenetic correlation between 0 (phylogenetic independence) and 1 (shared evolutionary history) and we implemented the mode covariance because values for each variables were on the same scale following transformation (\protect\hyperlink{ref-abdi2010}{Abdi \& Williams 2010}). Moreover, to corroborate that our imputation of missing values did not affect our results, we conducted a pPCA on the full dataset without missing values (see Supplementary Fig. S2). We found little difference between the explained variance with the imputed dataset (51.08\%) and the dataset without missing values (52.87\%). In addition, the loadings on each principal component had a similar contribution and correlation patterns, with the exception of plant height which showed slight variations between the imputed and non-imputed dataset. Finally, we conducted an additional phylogenetic informed principal component analysis for the subset of species with pollen and nectar quantity. For this, we included all quantitative traits considered in the main pPCA plus pollen grains and microlitres of nectar per flower.

\textbf{Phylogenetic signal of traits.} We calculated the phylogenetic signal of the different quantitative traits on the imputed dataset with the full set of species (N = 1,506) with the package \emph{phytools} (\protect\hyperlink{ref-revell2012}{Revell 2012}) version \emph{0.7-70} and we used Pagel's \(\lambda\) as a measurement of the phylogenetic signal. However, for pollen and nectar traits, phylogenetic signal was calculated only on the subset of species that had quantitative information for these traits (N = 755).

\textbf{Network analyses.} Analyses were conducted on the subset of 60 weighted networks sampled in a unique flowering season and site, which included 556 plant and 1,126 pollinator species. These networks were analysed in their qualitative (presence-absence) and quantitative (interaction frequency) form. First, we analysed the binary version of these weighted networks with presence-absence information that assumes equal weight across interactions. Second, we analysed the untransformed weighted networks with interaction frequency that accounts for the intensity of the interaction. Although floral visitors are not always pollinators and interaction frequency does not consider each pollinator species efficiency (\protect\hyperlink{ref-ballantyne2015}{Ballantyne \emph{et al.} 2015}), interaction frequency can provide valuable information of the contribution of floral visitors to pollination (\protect\hyperlink{ref-vazquez2005}{Vázquez \emph{et al.} 2005}, \protect\hyperlink{ref-vazquez2012}{2012}). In total, our network dataset (excluding meta-webs and non-weighted networks) included 2,256 interactions of bees with plants, 1,768 non-syrphid-Diptera interactions, 845 syrphids interactions, 437 Lepidoptera interactions, 432 Coleoptera interactions and 362 non-bee-Hymenoptera interactions. Sampling methods varied across networks but this was accounted for in analyses by considering them in the random effects of the modelling process. All analyses were conducted in R version \emph{4.0.3}.

\textbf{Visitation patterns.} We used Bayesian modelling (see below for details) to explore the effect of floral visitor groups and the main axes of trait variation (pPCA with imputed dataset) on both qualitative (presence/absence) and quantitative (visitation rate) floral interactions per plant species. For this, we divided floral visitors into six main guilds that differ in life form, behaviour and are likely to play a similar ecological role: (i) bees (Hymenoptera-Anthophila), (ii) non-bee-Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Diptera-Syrphidae), (iv) non-syrphid-Diptera (Diptera-non-Syrphidae), (v) Lepidoptera and (vi) Coleoptera. Moreover, because the guild of bees was the most represented group with 2,256 records and had the highest frequency of visits of all groups, we also explored the presence-absence of interaction and visitation rate of the main bee families (Andrenidae, Apidae, Colletidae, Halictidae and Megachilidae) on the trait space. In addition, we found that \emph{Apis mellifera} was the floral visitor with the largest proportion of records counted (7.55\% of the total). This finding is consistent with previous research showing that \emph{A. mellifera} was the most frequent floral visitor in a similar dataset of 80 plant-pollinator networks in natural ecosystems (\protect\hyperlink{ref-hung2018}{Hung \emph{et al.} 2018}). Hence, to control for the effect of \emph{A. mellifera} on the observed visitation patterns of bees, we conducted an analogous analysis with presence-absence of interaction and visitation rate excluding \emph{A. mellifera}. We found that \emph{A. mellifera}, was partly driving some of the observed trends on PC1 (Supplementary Fig. S3). However, we did not detect major differences on PC2 and PC3.

@@

\textbf{Plant species functional roles.} We investigated whether different quantitative traits determined plant species functional roles using Bayesian modelling and regression trees. For this, we selected simple and complementary species-level network metrics commonly applied in bipartite network studies (\protect\hyperlink{ref-dormann2008}{Dormann \emph{et al.} 2008}) with a straightforward ecological interpretation relevant to our research goals. The different plant species-level metrics were: (i) sum of visits per plant species; (ii) normalized degree, calculated as the number of links per plant species divided by the total possible number of partners; and (iii) specialization (d') (\protect\hyperlink{ref-bluthgen2006}{Blüthgen \emph{et al.} 2006}), which measures the deviation of an expected random choice of the available interaction partners and ranges between 0 (maximum generalization) and 1 (maximum specialization). Normalized degree and specialization were calculated with the \emph{specieslevel} function from the R package \emph{bipartite} (\protect\hyperlink{ref-dormann2008}{Dormann \emph{et al.} 2008}) (version \emph{2.15}).

@@

\textbf{Plant strategies.} The phylogenetically informed principal component analysis (pPCA) captured by the first two and three axes 51.8\% and 70.97\% of trait variation, respectively (Fig. 1 and Supplementary Fig. S5) and had a phylogenetic correlation (\(\lambda\)) of 0.76. The first principal component (PC1) represented 26.72\% of the trait variation and indicated a trade-off between flower number and flower size. We refer to this axis as the `flower number - flower size trade-off,' as already described in previous studies (\protect\hyperlink{ref-sargent2007}{Sargent \emph{et al.} 2007}; \protect\hyperlink{ref-kettle2011}{Kettle \emph{et al.} 2011}). Hence, one end of the spectrum comprised species with high investment in flower number and plant height but small flower size, short style length and low ovule number. The other end of this spectrum comprised species that were short in height and invested in large flowers, long styles, many ovules, but few flowers. The main contributing traits to PC1 were plant height, flower number, ovule number and flower size (loadings \textgreater{} \textbar0.5\textbar; Supplementary Table S3) but style length also contributed moderately to PC1 (loading = -0.33). The second principal component (PC2) represented 25.05\% of the trait variation and indicated a trade-off between low and high pollinator dependence. We refer to this axis as the `pollinator dependence trade-off.' The main driver of trait variation on PC2 was autonomous selfing (loading = 0.85) but the other traits (except ovule number) also made moderate contributions (loadings from 0.27 to 0.4; Supplementary Table S3). We found that high pollinator dependence was associated with larger and a higher number of flowers, greater plant height and longer styles. In contrast, species with high levels of autonomous selfing tended to have fewer and smaller flowers, had shorter styles and were shorter in height. Further, PC3 explained a considerable amount of trait variability (19.17\%) and the main contributors to this axis were style length (loading = -0.66) and the degree of autonomous selfing (loading = -0.51). The remaining traits, apart from ovule number, were moderately correlated to changes on PC3 (loadings from -0.23 to -0.46; Supplementary Table S3). Thus, because style length was correlated with all traits on PC3 and was the main driver of trait variation, we refer to this axis as the `style length trade-off.' Further, the pPCA with the subset of species that had nectar and pollen quantity data showed that nectar quantity (microlitres of nectar per flower) was positively associated with flower size, style length and ovule number (PC1, 23.40\%); and pollen quantity (pollen grains per flower) was positively correlated with flower number and plant height and negatively associated with autonomous selfing (PC2, 21.67\%; Supplementary Fig. S6). This pPCA explained similar variance with the first two principal components (45.07\%) and similar associations of traits despite some variability in the loadings (Supplementary Table S4).

@@

\textbf{Phylogenetic signal of traits.} We found a strong phylogenetic signal (\emph{P} \textless{} 0.01) in all quantitative traits (Supplementary Table S5). The traits that showed the highest phylogenetic signal were ovule number (\(\lambda\) = 1), pollen grains per flower (\(\lambda\) = 1) and plant height (\(\lambda\) = 0.96), followed by flower length (\(\lambda\) = 0.75), flower width (\(\lambda\) = 0.73), number of flowers per plant (\(\lambda\) = 0.69) and nectar concentration (\(\lambda\) = 0.65). The traits that showed a moderate phylogenetic signal were inflorescence width (\(\lambda\) = 0.57), style length (\(\lambda\) = 0.49) and autonomous selfing (\(\lambda\) = 0.34). Finally, microliters of nectar per flower showed the lowest phylogenetic signal of all traits (\(\lambda\) = 0.14).

\textbf{Visitation patterns.} The main axes of trait variation explained partly presence-absence interactions between plant and floral visitors (conditional \(R^{2}\) = 0.26; marginal \(R^{2}\) = 0.20) but little of the overall visitation rates (conditional \(R^{2}\) = 0.31; marginal \(R^{2}\) = 0.06). However, we found relevant trends across the different floral visitor guilds on both presence-absence and visitation interactions (Fig. 3). On the pollinator dependence trade-off, all floral visitor guilds interacted more frequently with plant species with higher pollinator dependence (PC2; Fig. 3b and Fig. 3e). For presence-absence interactions we found that all Diptera, Coleoptera and non-bee-Hymenoptera guilds interacted more frequently with plants with high flower number and small flowers (flower number - flower size trade-off, PC1; Fig. 3a) but bees and Lepidoptera interacted slightly more frequently with plant species with low flower number but large flowers. For presence-absence interactions on PC3 (style length trade-off; Fig. 3c), we found that bees interacted clearly more with plant species with long styles and high selfing and the rest of the guilds interacted slightly more with plant species with short styles and low selfing. In addition, all guilds other than Syrphids and Lepidoptera (i.e., all Hymenoptera, non-syrphid-Diptera and Coleoptera) showed greater visitation rates on species with small numerous flowers (PC1; Fig. 3d). On the style length trade-off, bees, Lepidoptera and non-bee-Hymenoptera showed greater visitation rates on plant species with larger styles and higher levels of selfing; while syrphids, non-syrphid-Diptera and Coleoptera showed higher visitation rates on species with shorter styles and lower selfing (Fig. 3f).

@@

\textbf{Plant species functional roles.} The variance of the different plant species-level network metrics was poorly explained by the three main axes of trait variation (Supplementary Fig. S9; interaction frequency \textasciitilde{} PCs, conditional \(R^{2}\) = 0.11, marginal \(R^{2}\) = 0.02; normalized degree \textasciitilde{} PCs, conditional \(R^{2}\) = 0.24, marginal \(R^{2}\) = 0.02; and, specialization \textasciitilde{} PCs, conditional \(R^{2}\) = 0.37, marginal\(R^{2}\) = 0.03). Overall, the most notable trends were found on PC1 and PC3 for interaction frequency and specialization. On the flower number - flower size trade-off (PC1), interaction frequency was higher for plant species with more flowers but was lower for plant species with larger flowers (Supplementary Fig. S9a). On PC1, specialization showed the opposite trend (Supplementary Fig. S9g). On the style length trade-off (PC3), interaction frequency was lower for plants with shorter styles and lower autonomous selfing and higher for species with longer styles and higher autonomous selfing (Supplementary Fig. S9c). Again, specialization showed the opposite trend to interaction frequency (Supplementary Fig. S9i).

>>>>

\textbf{Plant-pollinator network studies}

We selected 28 studies from 18 different countries that constituted a total of 64 plant-pollinator networks. These studies recorded plant-pollinator interactions in natural systems and were selected so that we had broad geographical representation. Although these studies differ in sampling effort and methodology, all studies provided information about plant-pollinator interactions (weighted and non-weighted), which we used to build a database of plant species that are likely to be animal pollinated. Many of these networks are freely available either as published studies (\protect\hyperlink{ref-olesen2007}{Olesen \emph{et al.} 2007}; \protect\hyperlink{ref-fortuna2010}{Fortuna \emph{et al.} 2010}; \protect\hyperlink{ref-carvalheiro2014}{Carvalheiro \emph{et al.} 2014}) or available in online archives (e.g., The Web of Life, \protect\hyperlink{ref-fortuna2010}{Fortuna \emph{et al.} 2010}; and Mangal, \protect\hyperlink{ref-poisot2016}{Poisot \emph{et al.} 2016}). In total, our network dataset (see Table S1) constituted 60 weighted (interaction frequency) and 4 unweighted (presence/absence of the interaction) networks, each sampled at a unique location and year, as well as eight meta-webs where interactions were pooled across several locations and multiple years.

\textbf{Taxonomy of plants and pollinators}

All species names, genera, families and orders were retrieved and standardized from the taxonomy data sources NCBI (\url{https://www.ncbi.nlm.nih.gov/taxonomy}) for plants and ITIS (\url{https://www.itis.gov/}) for pollinators, using the R package \emph{taxize} (\protect\hyperlink{ref-chamberlain2020}{Chamberlain \emph{et al.} n.d.b}) version \emph{0.9.99}. We filled the `not found' searches manually using \url{http://www.theplantlist.org/} and \url{http://www.mobot.org/} for plants and \url{http://www.catalogueoflife.org/} for floral visitors.

\textbf{Functional traits}

We selected a total of 19 different functional traits that included both reproductive and vegetative traits (see Table 1 and Supplementary Information). From these, 16 were reproductive traits (13 floral and 3 reproductive biology traits) and were selected based on their relevance to plant reproduction and data availability. The 3 remaining traits, were vegetative traits that are commonly used to characterize the global spectrum of plant form and function and represent the fast-slow continuum of trait variation (e.g., short-lived versus perennial species). For each plant species, we undertook an extensive literature and online search across a wide range of resources (plant databases, online floras, books, journals and images). From a total of 30,120 cells (20 columns × 1,506 species) we were able to fill 24,341 cells (80.8\% of the dataset, see Fig. S1 for missing values information for each trait).

\textbf{Phylogenetic Distance}

We calculated the phylogenetic distance between different plant species using the function \emph{get\_tree} from the package \emph{rtrees} (\url{https://github.com/daijiang/rtrees}), which downloads phylogenetic distances from the extended R implementation of the Open Tree of Life (\protect\hyperlink{ref-smith2018}{Smith \& Brown 2018}; \protect\hyperlink{ref-jin2019}{Jin \& Qian 2019}).

@@

\textbf{Data Imputation}

Trait missing values were imputed with the function \emph{missForest} (\protect\hyperlink{ref-stekhoven2012}{Stekhoven \& Bühlmann 2012}) which allows imputation of data sets with continuous and categorical variables. We accounted for the phylogenetic distance among species on the imputation process by including the eigenvectors of a principal component analysis of the phylogenetic distance (PCoA) which has been shown to improve the performance of \emph{missForest} (\protect\hyperlink{ref-penone2014}{Penone \emph{et al.} 2014}). To extract the eigenvectors, we used the function \emph{PVRdecomp} from the package \emph{PVR} (\protect\hyperlink{ref-santos2018}{Chamberlain \emph{et al.} n.d.a}) based on a previous conceptual framework that considers phylogenetic eigenvectors (\protect\hyperlink{ref-diniz-filho2012}{Diniz-Filho \emph{et al.} 2012}). Although the variable of autonomous selfing had a high percentage of missing values (68\%), we were able to solve this by back transforming the qualitative column of autonomous selfing to numerical. The categories of `none,' `low,' `medium' and `high' were converted to representative percentages of each category 0\%, 13\%, 50.5\% and 88\% respectively. This reduced the percentage of missing values for this column from 68\% to 35\% and allowed the imputation of this variable. However, we were unable to include nectar and pollen traits on the imputation process because of the high percentage of missing values (Supplementary Fig. S1). Hence, the imputed dataset had 1,506 species, seven categorical and eight numerical variables and 5.79\% of missing values. Further, we conducted an additional imputation process on the subset of species with data for pollen per flower and microliters of nectar. This subset comprised 755 species, 8.01\% missing values and all traits but milligrams of nectar (\textasciitilde50\% of missing values) were included in the imputation process.

\textbf{Plant strategies}

We explored the trade-offs between different quantitative plant functional traits with a phylogenetically informed Principal Component Analysis (pPCA). We did not include the quantitative variables of flower length and inflorescence width because they were highly and moderately correlated to flower width respectively (Pearson's correlation = 0.72, \emph{P} \textless{} 0.01 and Pearson's correlation = 0.36, \emph{P} \textless{} 0.01), and thus we avoided overemphasizing flower size on the spectrum of trait variation. Although qualitative traits were not included in the dimensionality reduction analysis, we also investigated the association of the different qualitative traits with the main axes of trait variation. Prior to the analyses, we excluded outliers and standardized the data. Due to the high sensitivity of dimensionality reduction to outliers, we excluded values within the 2.5th--97.5th percentile range (\protect\hyperlink{ref-legendre2012}{Legendre \& Legendre 2012}), and thus our final dataset had 1,236 species. Then, we log transformed the variables to reduce the influence of outliers and z-transformed (X= 0, SD=1) so that all variables were within the same numerical range. We performed the pPCA using the function \emph{phyl.pca} from the package \emph{phytools} (\protect\hyperlink{ref-revell2012}{Revell 2012}) (version \emph{0.7-70}) with the method lambda (\(\lambda\)) that calculates the phylogenetic correlation between 0 (phylogenetic independence) and 1 (shared evolutionary history) and we implemented the mode covariance because values for each variables were on the same scale following transformation (\protect\hyperlink{ref-abdi2010}{Abdi \& Williams 2010}). Moreover, to corroborate that our imputation of missing values did not affect our results, we conducted a pPCA on the full dataset without missing values (see Supplementary Fig. S2). We found little difference between the explained variance with the imputed dataset (51.08\%) and the dataset without missing values (52.87\%). In addition, the loadings on each principal component had a similar contribution and correlation patterns, with the exception of plant height which showed slight variations between the imputed and non-imputed dataset. Finally, we conducted an additional phylogenetic informed principal component analysis for the subset of species with pollen and nectar quantity. For this, we included all quantitative traits considered in the main pPCA plus pollen grains and microlitres of nectar per flower.

\textbf{Phylogenetic signal of traits}

We calculated the phylogenetic signal of the different quantitative traits on the imputed dataset with the full set of species (N = 1,506) with the package \emph{phytools} (\protect\hyperlink{ref-revell2012}{Revell 2012}) version \emph{0.7-70} and we used Pagel's \(\lambda\) as a measurement of the phylogenetic signal. However, for pollen and nectar traits, phylogenetic signal was calculated only on the subset of species that had quantitative information for these traits (N = 755).

\textbf{Network analyses}

Analyses were conducted on the subset of 60 weighted networks sampled in a unique flowering season and site, which included 556 plant and 1,126 pollinator species. These networks were analysed in their qualitative (presence-absence) and quantitative (interaction frequency) form. First, we analysed the binary version of these weighted networks with presence-absence information that assumes equal weight across interactions. Second, we analysed the untransformed weighted networks with interaction frequency that accounts for the intensity of the interaction. Although floral visitors are not always pollinators and interaction frequency does not consider each pollinator species efficiency (\protect\hyperlink{ref-ballantyne2015}{Ballantyne \emph{et al.} 2015}), interaction frequency can provide valuable information of the contribution of floral visitors to pollination (\protect\hyperlink{ref-vazquez2005}{Vázquez \emph{et al.} 2005}, \protect\hyperlink{ref-vazquez2012}{2012}). In total, our network dataset (excluding meta-webs and non-weighted networks) included 2,256 interactions of bees with plants, 1,768 non-syrphid-Diptera interactions, 845 syrphids interactions, 437 Lepidoptera interactions, 432 Coleoptera interactions and 362 non-bee-Hymenoptera interactions. Sampling methods varied across networks but this was accounted for in analyses by considering them in the random effects of the modelling process. All analyses were conducted in R version \emph{4.0.3}.

\textbf{Visitation patterns}

We used Bayesian modelling (see below for details) to explore the effect of floral visitor groups and the main axes of trait variation (pPCA with imputed dataset) on both qualitative (presence/absence) and quantitative (visitation rate) floral interactions per plant species. For this, we divided floral visitors into six main guilds that differ in life form, behaviour and are likely to play a similar ecological role: (i) bees (Hymenoptera-Anthophila), (ii) non-bee-Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Diptera-Syrphidae), (iv) non-syrphid-Diptera (Diptera-non-Syrphidae), (v) Lepidoptera and (vi) Coleoptera. Moreover, because the guild of bees was the most represented group with 2,256 records and had the highest frequency of visits of all groups, we also explored the presence-absence of interaction and visitation rate of the main bee families (Andrenidae, Apidae, Colletidae, Halictidae and Megachilidae) on the trait space. In addition, we found that \emph{Apis mellifera} was the floral visitor with the largest proportion of records counted (7.55\% of the total). This finding is consistent with previous research showing that \emph{A. mellifera} was the most frequent floral visitor in a similar dataset of 80 plant-pollinator networks in natural ecosystems (\protect\hyperlink{ref-hung2018}{Hung \emph{et al.} 2018}). Hence, to control for the effect of \emph{A. mellifera} on the observed visitation patterns of bees, we conducted an analogous analysis with presence-absence of interaction and visitation rate excluding \emph{A. mellifera}. We found that \emph{A. mellifera}, was partly driving some of the observed trends on PC1 (Supplementary Fig. S3). However, we did not detect major differences on PC2 and PC3.

@@

\textbf{Plant species functional roles}

We investigated whether different quantitative traits determined plant species functional roles using Bayesian modelling and regression trees. For this, we selected simple and complementary species-level network metrics commonly applied in bipartite network studies (\protect\hyperlink{ref-dormann2008}{Dormann \emph{et al.} 2008}) with a straightforward ecological interpretation relevant to our research goals. The different plant species-level metrics were: (i) sum of visits per plant species; (ii) normalized degree, calculated as the number of links per plant species divided by the total possible number of partners; and (iii) specialization (d') (\protect\hyperlink{ref-bluthgen2006}{Blüthgen \emph{et al.} 2006}), which measures the deviation of an expected random choice of the available interaction partners and ranges between 0 (maximum generalization) and 1 (maximum specialization). Normalized degree and specialization were calculated with the \emph{specieslevel} function from the R package \emph{bipartite} (\protect\hyperlink{ref-dormann2008}{Dormann \emph{et al.} 2008}) (version \emph{2.15}).

@@

\textbf{Plant strategies}

The phylogenetically informed principal component analysis (pPCA) captured by the first two and three axes 51.8\% and 70.97\% of trait variation, respectively (Fig. 1 and Supplementary Fig. S5) and had a phylogenetic correlation (\(\lambda\)) of 0.76. The first principal component (PC1) represented 26.72\% of the trait variation and indicated a trade-off between flower number and flower size. We refer to this axis as the `flower number - flower size trade-off,' as already described in previous studies (\protect\hyperlink{ref-sargent2007}{Sargent \emph{et al.} 2007}; \protect\hyperlink{ref-kettle2011}{Kettle \emph{et al.} 2011}). Hence, one end of the spectrum comprised species with high investment in flower number and plant height but small flower size, short style length and low ovule number. The other end of this spectrum comprised species that were short in height and invested in large flowers, long styles, many ovules, but few flowers. The main contributing traits to PC1 were plant height, flower number, ovule number and flower size (loadings \textgreater{} \textbar0.5\textbar; Supplementary Table S3) but style length also contributed moderately to PC1 (loading = -0.33). The second principal component (PC2) represented 25.05\% of the trait variation and indicated a trade-off between low and high pollinator dependence. We refer to this axis as the `pollinator dependence trade-off.' The main driver of trait variation on PC2 was autonomous selfing (loading = 0.85) but the other traits (except ovule number) also made moderate contributions (loadings from 0.27 to 0.4; Supplementary Table S3). We found that high pollinator dependence was associated with larger and a higher number of flowers, greater plant height and longer styles. In contrast, species with high levels of autonomous selfing tended to have fewer and smaller flowers, had shorter styles and were shorter in height. Further, PC3 explained a considerable amount of trait variability (19.17\%) and the main contributors to this axis were style length (loading = -0.66) and the degree of autonomous selfing (loading = -0.51). The remaining traits, apart from ovule number, were moderately correlated to changes on PC3 (loadings from -0.23 to -0.46; Supplementary Table S3). Thus, because style length was correlated with all traits on PC3 and was the main driver of trait variation, we refer to this axis as the `style length trade-off.' Further, the pPCA with the subset of species that had nectar and pollen quantity data showed that nectar quantity (microlitres of nectar per flower) was positively associated with flower size, style length and ovule number (PC1, 23.40\%); and pollen quantity (pollen grains per flower) was positively correlated with flower number and plant height and negatively associated with autonomous selfing (PC2, 21.67\%; Supplementary Fig. S6). This pPCA explained similar variance with the first two principal components (45.07\%) and similar associations of traits despite some variability in the loadings (Supplementary Table S4).

@@

\textbf{Phylogenetic signal of traits}

We found a strong phylogenetic signal (\emph{P} \textless{} 0.01) in all quantitative traits (Supplementary Table S5). The traits that showed the highest phylogenetic signal were ovule number (\(\lambda\) = 1), pollen grains per flower (\(\lambda\) = 1) and plant height (\(\lambda\) = 0.96), followed by flower length (\(\lambda\) = 0.75), flower width (\(\lambda\) = 0.73), number of flowers per plant (\(\lambda\) = 0.69) and nectar concentration (\(\lambda\) = 0.65). The traits that showed a moderate phylogenetic signal were inflorescence width (\(\lambda\) = 0.57), style length (\(\lambda\) = 0.49) and autonomous selfing (\(\lambda\) = 0.34). Finally, microliters of nectar per flower showed the lowest phylogenetic signal of all traits (\(\lambda\) = 0.14).

\textbf{Visitation patterns}

The main axes of trait variation explained partly presence-absence interactions between plant and floral visitors (conditional \(R^{2}\) = 0.26; marginal \(R^{2}\) = 0.20) but little of the overall visitation rates (conditional \(R^{2}\) = 0.31; marginal \(R^{2}\) = 0.06). However, we found relevant trends across the different floral visitor guilds on both presence-absence and visitation interactions (Fig. 3). On the pollinator dependence trade-off, all floral visitor guilds interacted more frequently with plant species with higher pollinator dependence (PC2; Fig. 3b and Fig. 3e). For presence-absence interactions we found that all Diptera, Coleoptera and non-bee-Hymenoptera guilds interacted more frequently with plants with high flower number and small flowers (flower number - flower size trade-off, PC1; Fig. 3a) but bees and Lepidoptera interacted slightly more frequently with plant species with low flower number but large flowers. For presence-absence interactions on PC3 (style length trade-off; Fig. 3c), we found that bees interacted clearly more with plant species with long styles and high selfing and the rest of the guilds interacted slightly more with plant species with short styles and low selfing. In addition, all guilds other than Syrphids and Lepidoptera (i.e., all Hymenoptera, non-syrphid-Diptera and Coleoptera) showed greater visitation rates on species with small numerous flowers (PC1; Fig. 3d). On the style length trade-off, bees, Lepidoptera and non-bee-Hymenoptera showed greater visitation rates on plant species with larger styles and higher levels of selfing; while syrphids, non-syrphid-Diptera and Coleoptera showed higher visitation rates on species with shorter styles and lower selfing (Fig. 3f).

@@

\textbf{Plant species functional roles}

The variance of the different plant species-level network metrics was poorly explained by the three main axes of trait variation (Supplementary Fig. S9; interaction frequency \textasciitilde{} PCs, conditional \(R^{2}\) = 0.11, marginal \(R^{2}\) = 0.02; normalized degree \textasciitilde{} PCs, conditional \(R^{2}\) = 0.24, marginal \(R^{2}\) = 0.02; and, specialization \textasciitilde{} PCs, conditional \(R^{2}\) = 0.37, marginal\(R^{2}\) = 0.03). Overall, the most notable trends were found on PC1 and PC3 for interaction frequency and specialization. On the flower number - flower size trade-off (PC1), interaction frequency was higher for plant species with more flowers but was lower for plant species with larger flowers (Supplementary Fig. S9a). On PC1, specialization showed the opposite trend (Supplementary Fig. S9g). On the style length trade-off (PC3), interaction frequency was lower for plants with shorter styles and lower autonomous selfing and higher for species with longer styles and higher autonomous selfing (Supplementary Fig. S9c). Again, specialization showed the opposite trend to interaction frequency (Supplementary Fig. S9i).
